\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-06-16 Articles/2020-QuantumLogic/Source/quantumlogic.tex
QLogic: final pass from Andrea

\begin{abstract}
\center Notes on classical and quantum logic
\vspace{4mm}

\noindent \emph{Keywords}: Quantum Mechanics; Quantum Logic; Classical Logic; 
\end{abstract}
\vspace{5mm}
\clearpage

@@

As soon as Quantum Mechanics (QM) achieved a definite and coherent mathematical formulation, several physicists and philosophers claimed that its formal structure does not conform to the laws of classical propositional calculus, a belief that is widespread still to this day (cf.\ \cite{Giuntini:2002}, \cite{DallaChiara:2004} and \cite{deRonde:2016}). From the thirties, in fact, various modifications of Classical Logic (CL) were advanced in order to faithfully represent the physical content of quantum theory. For instance, in 1931 the Polish philosopher Zygmunt Zawirski proposed to apply \L ukasiewicz's three-valued logic to QM, starting from considerations about the wave-particle duality and Heisenberg's uncertainty principle. In virtue of the latter, moreover, the logical empiricists Moritz Schlick and Philipp Frank claimed that the conjunction of two or more statements may have no meaning in quantum theory, concluding that the classical conjunction is not always valid in the context of quantum physics, as recalled in \cite{Carnap:1966}. In their view, a typical example of invalid conjunction is one attributing a definite value for both position and momentum to a single quantum particle at a given time $t$. Furthermore, in 1933 Fritz Zwicky suggested that QM rejects the law of excluded-middle.\footnote{For historical details on quantum logic see \cite{Jammer:1974}.} However, it is generally recognized that 
the founding text on Quantum Logic (QL) is Birkhoff and von Neumann's essay ``\emph{The Logic of Quantum Mechanics}'' (\cite{vonNeumann:1936}), where the authors provided the standard account of the propositional calculus obeyed by quantum propositions.\footnote{It is worth noting that von Neumann in his treatise \emph{Mathematische Grundlagen der Quantenmechanik} published in 1932 anticipated that from the algebraic structure of quantum theory it would have been possible to formulate a new propositional calculus. These ideas would have been fully expressed of his successive collaboration with Garrett Birkhoff.}

A few years after the publication of this paper, the research in quantum logic had a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct formal structures to represent the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now accepted among experts; for details see e.g.\ by \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it is an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and Classical Mechanics (CM), following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First order and higher order logics will not be discussed here as well as the logic of relativistic quantum theories.}

Recently one of us faced the question of whether quantum physics necessarily implies a revision of classical logic (cf.\ \cite{Oldofredi:2020}). Alternatively stated, it was asked if classical logic must be abandoned in the quantum realm, or there are quantum theories compatible with a classical propositional calculus. Analyzing Bohmian Mechanics, a well-known interpretation of QM, it has been showed that quantum physics does not necessarily involves a departure from classical logic. More precisely, it has been argued that in the context of this interpretation of QM not only the logical connectives retain their classical meanings---in particular the negation and the disjunction do follow classical rules---but also the distributivity law holds, contrary to the case of standard QM. This fact is a consequence of the clear ontology of the theory and its explanation of quantum measurements.

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, here we have a two-fold aim: in the first place, we will analyze the main motivations and justifications for the introduction of quantum logic arguing that they do not entail the failure of CL in the quantum context. In the second place, we show that from the mathematical structures on which QM is built, it is possible to provide a distributive lattice to quantum propositions, contrary to the received view about QL. Such a distributive lattice of quantum statements, moreover, will allow us to take into account also assertions concerning expectation values, which are generally excluded in the context of the standard approach to quantum logic. Finally, we will argue that it is possible to demonstrate that classical logical propositions can form a non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: Section \ref{QL} provides a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part), while in Section \ref{Motivations} we review the main arguments that have been given in literature to introduce a quantum logical calculus. Here we explain  that they do not entail the failure of classical logic in QM. In Section \ref{Math} we show not only that quantum logical propositions generate a distributive lattice, but also that important sentences of quantum theory---i.e.\ those concerning expectations values---are excluded by standard QL.\ In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

Since part of the arguments in favor of QL stem from a claimed unsuitability of classical logic in the context of quantum mechanics, let us say a few words about classical logic, classical mechanics and their purported relationship.  It is a well known result that classical propositional logic is equivalent to a Boolean algebra where propositions (atomic and complex) have truth values ``true'' and ``false'' (or $\top$ and $\bot$), and the main operations among them are conjunction ``$\wedge$'', disjunction ``$\vee$'' and negation ``$\neg$'', the only unary operation. From these logical connectives one can introduce secondary operations, as for instance the material implication ``$\longrightarrow$'', the exclusive or ``$\oplus$'' and logical equivalence ``$\longleftrightarrow$''. Furthermore, it is important for our discussion to underline that the laws of propositional logic include commutativity, associativity, identity and distributivity with respect to $\wedge$ and $\vee$. It is also well established that the algebraic structure underlying classical mechanics is commutative, distributive and associative. The standard view in QL is that classical logic is applicable to classical mechanics because of the commutative algebraic structure.

@@

Once the previous stance is taken, one necessarily concludes that classical propositional calculus is not appropriate to represent the logic obeyed by quantum propositions, since quantum theory does not share the same algebraic structure of classical mechanics. Referring to this, \cite{Giuntini:2002} claim that distributivity, a fundamental property of classical logic, fails in quantum logic for two main reasons:
\begin{itemize}
	\item the non-commutativity of quantum observables,
	\item the peculiar behavior of the quantum disjunction---the second major deviation from classical logic---which may be true even though neither of the members is true in virtue of the linearity of the Schr\"odinger Equation (SE). 
\end{itemize}
\noindent Let us then have a closer look at these motivations for the failure of the distributivity law, and more generally of CL.

@@

Let us consider a $1/2-$spin particle, such as an electron. From QM we know that there are only two possible spin states in which a particle can be found along each axis after a spin measurement, namely either in the up or in the down state. Furthermore, since spin operators along different axes are incompatible and do not commute---i.e.\ $[S_x, S_y]\neq 0, [S_x, S_z]\neq 0, [S_y, S_z]\neq0$---they represent incompatible observables that cannot be simultaneously measured. Let us then consider the following scenario: an electron is prepared with spin along the $x$ direction and subsequently the spin along the $y$ axis is measured. Let the propositions $q$, $p$ and $r$ be defined as follows:
\begin{description}
	\item $p$: ``the electron has $x$-spin up'', 
	\item $q$: ``the electron has $y$-spin up'',
	\item $r$: ``the electron has $y$-spin down''.
\end{description}
As we said, there are only two states in which a particle can be found along an axis. Therefore the proposition $q \vee r$ ``must be true'' (\cite{Giuntini:2002}, pp.\ 133-134). However, because the different directions are incompatible, if $p$ is true then both $q$ and $r$ must be false. The result is, as noted before, that the quantum disjunction is true even if neither of its members is true.

This behavior in turn breaks the law of distributivity which would state:
\begin{align}
	\label{DL}
	p\wedge(q\vee r)\longleftrightarrow(p\wedge q)\vee(p\wedge r).
\end{align}
\noindent Since we assume that $p$ is true and we know that the disjunction $(q\vee r)$ is also true, whereas $(p\wedge q)$ and $(p\wedge r)$ are both false in virtue of the incompatibility between the spin operators among different axes. This fact implies that the right hand side of the distributive law asserts a false statement, i.e.\ $(p\wedge q)\vee(p\wedge r)=\bot$. Consequently, we have 
\begin{align}
	\label{contradiction}
	p\wedge(q\vee r)=\top\longleftrightarrow(p\wedge q)\vee(p\wedge r)=\bot. 
\end{align}
Therefore the distributive law fails in the context of quantum theory.\footnote{In the context of quantum logic, it is usually believed that distributivity must be replaced by a weaker law: $(p\wedge q)\vee(p\wedge r)\longrightarrow(p\wedge(q\vee r)).$}

@@

Their analysis begins by defining quantum logical propositions as \emph{experimental propositions}, i.e.\ statements affirming that a certain observable $A$ (or a set of observables) measured on a quantum system $s$ has a given value $a_i$ (or a sequence of values). More precisely, the authors stress that both in classical and quantum mechanics observations of physical systems are given by readings of experimental outcomes ($x_1, \dots, x_n$) of \emph{compatible} measurements ($\mu_1, \dots, \mu_n$). The values ($x_1, \dots, x_n$) are elements of what the authors called the ($x_1, \dots, x_n$)-space, i.e.\ the ``observation-space'' of the system in question, whose elements are all the possible combinations of results of the compatible measurements ($\mu_1, \dots, \mu_n$). Hence, the actual values ($x_1, \dots, x_n$) form a subset of such a space. Birkhoff and von Neumann, then, defined the experimental propositions concerning a physical system as the subsets of the observation space associated with it.

Secondly, the authors underlined that in classical and quantum mechanics the states of physical systems are mathematically represented by points in their state spaces---phase space for the classical case, and Hilbert space $\mathcal{H}$ for the quantum case---which provide the maximal information concerning the system. A point in phase space corresponds to the specification of the position and momentum variables of a certain classical system, whereas in QM the points of $\mathcal{H}$ correspond to wave functions. The authors, then, find a connection between subsets of the observation-space of a system and subsets of its Hilbert space, specifying that quantum experimental propositions are mathematically represented by a closed linear subspace of $\mathcal{H}$; this step is crucial in order to obtain the quantum propositional calculus.\ Alternatively, we can say that quantum mechanical operators correspond to propositions with ``yes/no'' (``true/false'') outcome in a logical system\footnote{Cf.\ also \cite{David:2015}, p.\ 78, where we read that ``An orthogonal projector $\textbf{P}$ onto a linear subspace $P\subset\mathcal{H}$ is indeed the operator associated to an observable that can take only the values 1 (and always 1 if the state $\psi\in P$ is in the subspace $P$) or 0 (and always 0 if the state $\psi\in P^{\perp}$ belongs to the orthogonal subspace to $P$).\ Thus we can consider that measuring the observable $\textbf{P}$ is equivalent to perform a test on the system, or to check the validity of a logical proposition $p$ on the system''. [TODO: check quote as it is grammatically incorrect]} as underlined by Svozi:
\begin{quote}
Any closed linear subspace of---or, equivalently, any projection operator on---a Hilbert space corresponds to an elementary
proposition. The elementary \emph{true--false} proposition can in English be spelled out explicitly as
\begin{quote}
``The physical system has a property corresponding to the associated closed linear subspace'' (\cite{Svozi:1999}, p.\ 1).
\end{quote}
\end{quote}

@@

In this section we will review the examples given in the previous one and see that the failure of distributivity stems from the ambiguity of the propositions and, once that is solved, standard classical rules can be applied. More precisely, we are going to argue that (i) it is not necessarily the case that a quantum disjunction must be true when neither of its members is, and (ii) that \eqref{DL} can be retained in QM, since the arguments seen above---usually meant to show its failure---contain serious flaws.

@@

As noted in the previous section, in classical mechanics one generally considers physical systems as objects having well-defined values for inherent, dynamical and relational properties which do not depend on any observation---meaning that measuring a magnitude on a classical particle will reveal a pre-existing value of the observed quantity. Thus, statements about the ontic state of classical systems before or after a measurement can be conflated without any dangerous metaphysical consequence.

However, in the context of QM this move is not allowed. The effect of the measurement on the state of the system, however one may choose to interpret them, cannot be disregarded. A statement about a system before a measurement, about its preparation\footnote{In operational quantum theory the preparations are a set of instructions that a certain agent has to follow in order to prepare a quantum system in a certain state.}, is different from a statement about the same system after its measurement, about the measurement outcome. In fact, it is only through repeated measurements that we can infer anything about the state of the system before the measurement. Thus, propositions like ``the electron has $y$-spin up'', ``the velocity of the particle is $v$'', ``the position of the particle is $x$'' are ambiguous as they do not spell out whether they are about preparation or measurement outcome. Once that is taken into account, the purported failure of classical logic disappears.

Let us then redefine the propositions as follows:
\begin{description}
	\item $p_i$: ``before the measurement, the electron $x$-spin is up'', 
	\item $p_o$: ``after the measurement, the electron $x$-spin is up'', 
	\item $q_i$: ``before the measurement, the electron $y$-spin is up'',
	\item $q_o$: ``after the measurement, the electron $y$-spin is up'',
	\item $r_i$: ``before the measurement, the electron $y$-spin is down'',
	\item $r_o$: ``after the measurement, the electron $y$-spin is down''.
\end{description}
\noindent Given the setup of the example, after the measurement either $q_o$ or $r_o$ must be true, which means $q_o \vee r_o = \top$. This reflects the simple fact that after a measurement of spin along the $y$ axis the state must be one of the two, which is true only after the measurement. The key is that $q_i$ and $r_i$ are not the same statements as $q_o$ and $r_o$, therefore it is not necessarily true that $q_o \vee r_o = q_i \vee r_i$. This unwarranted generalization is what causes the problem.

\tikzstyle{vecArrow} = [thick, decoration={markings,mark=at position
	1 with {\arrow[semithick]{open triangle 60}}},
double distance=1.4pt, shorten >= 5.5pt,
preaction = {decorate},
postaction = {draw,line width=1.4pt, white,shorten >= 4.5pt}]
\tikzstyle{innerWhite} = [semithick, white,line width=1.4pt, shorten >= 4.5pt]

@@

		%	\draw \setA;
		%	\begin{scope}
		%		\clip \setA;
		%		\draw (-5, 4) .. controls (0, 2) and (0, -2) .. (-5, -4);
		%		\draw (2, 4) .. controls (2, 2) and (0, 1) .. (-1.46, 1);
		%		\draw (1, 1.9) .. controls (1.5, 1) and (2, -2.5) .. (0.5, -4);
		%		\draw (1.49, -1)  .. controls (2, -0.3) and (4, 0.5) ..  (5, 0);
		%	\end{scope}
		%	\node at (-4, -4.5) {$A$};
		%	\node at (-3.2, 0) {$P_1$};
		%	\node at (-0.4, 2.6) {$P_2$};
		%	\node at (-0.2, -1.2) {$P_3$};
		%	\node at (3, -1.7) {$P_4$};
		%	\node at (3, 1.5) {$P_5$};
	\end{tikzpicture} 
}
\end{center}

>>>>

\begin{abstract}
It is generally believed that quantum mechanics entails a rejection of classical logic. The main arguments to support this claim come from the non-commutativity of quantum observables, that allegedly would generate a non-distributive lattice of propositions, and from quantum superpositions, which would entail new rules for quantum disjunctions. In this essay we show that such arguments contain serious flaws: on the one hand, it is possible to keep the classical distributivity law in the context of quantum theory once the actual meaning of quantum propositions is unambiguously defined and properly taken into account. On the other hand, we argue that the existence of quantum superpositions does not lead to the definition of a new rule for logical disjunctions. Furthermore, contrary to the received view in quantum logic, we explain that quantum mechanics can generate a distributive lattice of propositions. Finally, the remarkable formal similarities between the logical structures of classical and quantum mechanics are highlighted and discussed. 
\vspace{4mm}

\noindent \emph{Keywords}: Quantum Mechanics; Quantum Logic; Classical Logic; Borel Algebra; Distributivity; Distributive Lattice
\end{abstract}
\vspace{5mm}
\clearpage

@@

As soon as Quantum Mechanics (QM) achieved a definite and coherent mathematical formulation, several physicists and philosophers claimed that its formal structure does not conform to the laws of classical propositional calculus, a belief that is widespread still to this day (cf.\ \cite{Giuntini:2002}, \cite{DallaChiara:2004} and \cite{deRonde:2016}). From the thirties, indeed, various modifications of Classical Logic (CL) were advanced in order to faithfully represent the physical content of quantum theory. For instance, in 1931 the Polish philosopher Zygmunt Zawirski proposed to apply \L ukasiewicz's three-valued logic to QM, starting from considerations about the wave-particle duality and Heisenberg's uncertainty principle. In virtue of the latter, moreover, the logical empiricists Moritz Schlick and Philipp Frank claimed that the conjunction of two or more statements may have no meaning in quantum theory, concluding that the classical conjunction is not always valid in the context of quantum physics, as recalled in \cite{Carnap:1966}. In their view, a typical example of invalid conjunction is one attributing a definite value for both position and momentum to a single quantum particle at a given time $t$. Furthermore, in 1933 Fritz Zwicky suggested that QM rejects the law of excluded-middle.\footnote{For historical details on quantum logic see \cite{Jammer:1974}.} However, it is generally recognized that the founding text of Quantum Logic (QL) is Birkhoff and von Neumann's essay ``\emph{The Logic of Quantum Mechanics}'' (\cite{vonNeumann:1936}), where the authors provided the standard account of the propositional calculus obeyed by quantum propositions.\footnote{It is worth noting that von Neumann in his treatise \emph{Mathematische Grundlagen der Quantenmechanik} published in 1932 anticipated that from the algebraic structure of quantum theory it would have been possible to formulate a new propositional calculus. These ideas would have been fully expressed of his successive collaboration with Garrett Birkhoff.}

A few years after the publication of this paper, the research in quantum logic saw a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct logical structures to understand and describe the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now accepted among experts; for details see e.g.\ by \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it is an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and Classical Mechanics (CM), following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First order and higher order logics will not be discussed here as well as the logic of relativistic quantum theories.}

Referring to this, one of us recently faced the question of whether quantum physics implies a revision of classical logic (cf.\ \cite{Oldofredi:2020}). Alternatively stated, it was asked if classical logic must be inevitably abandoned in the quantum realm, or if there are quantum theories compatible with classical propositional calculus. Analyzing Bohmian mechanics, a well-known interpretation of QM, it has been showed that quantum physics does not necessarily involves a departure from classical logic. More precisely, it has been argued that in this interpretation of QM not only the logical connectives retain their classical meanings---in particular the negation and the disjunction do follow classical rules---but also the distributivity law holds. This fact is a consequence of the particle ontology of the theory and its explanation of quantum measurements.

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics itself necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, this essay has a two-fold aim: in the first place, we analyze the main motivations and justifications for the introduction of quantum logic, claiming that such arguments do not really entail the failure of CL in the quantum context because they contain serious flaws. In the second place, we show that from the mathematical structures on which QM is built, it is possible to provide a distributive lattice of quantum propositions, contrary to the received view in QL. Such a distributive lattice of quantum statements, moreover, allows us to take into account assertions concerning expectation values, which are generally excluded in the standard approach to quantum logic. Finally, we argue that it is possible to demonstrate that classical logical propositions can form a non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: in Section \ref{QL} we review the main arguments that have been advanced in literature to introduce a quantum logical calculus, here we also provide a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part). Section \ref{Motivations} explains why the main examples used to show the failure of classical logic in QM contain important flaws and ambiguities; moreover, we argue that the classical distributive law can be maintained in quantum mechanics, once such ambiguities are removed. In Section \ref{Math}, contrary to the received view in QL, we show not only that quantum logical propositions can generate a distributive lattice, but also that important sentences of quantum theory---i.e.\ those concerning expectations values---are excluded by standard quantum logic.\ In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

Given that a significant part of the arguments in favor of QL stem from a claimed unsuitability of classical logic in the context of quantum mechanics, let us say a few words about classical logic, classical mechanics and their purported relationship.  It is a well known result that classical propositional logic is equivalent to a Boolean algebra where propositions (atomic and complex) have truth values ``true'' and ``false'' (or $\top$ and $\bot$), and the main operations among them are conjunction ``$\wedge$'', disjunction ``$\vee$'' and negation ``$\neg$'', the only unary operation. From these logical connectives one can introduce secondary operations, as for instance the material implication ``$\longrightarrow$'', the exclusive or ``$\oplus$'' and logical equivalence ``$\longleftrightarrow$''. Furthermore, it is important for our discussion to underline that the laws of propositional logic include commutativity, associativity, identity and distributivity with respect to $\wedge$ and $\vee$. It is also well established that the algebraic structure underlying classical mechanics is commutative, distributive and associative. The standard view in QL is that classical logic is applicable to classical mechanics because of its commutative algebraic structure.

@@

Once the previous stance is taken, one necessarily concludes that classical propositional calculus is not appropriate to represent the logic obeyed by quantum propositions, since quantum theory does not share the same algebraic structure of classical mechanics. Referring to this, \cite{Giuntini:2002} claim that distributivity, a fundamental property of classical logic, fails in quantum logic for two main reasons:
\begin{itemize}
	\item the non-commutativity of quantum observables,
	\item the peculiar behavior of the quantum disjunction---another major deviation from classical logic---which may be true even though neither of the members is true in virtue of the linearity of the Schr\"odinger Equation (SE). 
\end{itemize}
\noindent Let us then have a closer look at these motivations for the failure of the distributivity law, and more generally of CL, in the quantum domain.

@@

Let us consider a $1/2-$spin particle, such as an electron. From QM we know that there are only two possible spin states in which a particle can be found along each axis after a spin measurement, namely either in the up or in the down state. Furthermore, since spin operators along different axes are incompatible and do not commute---i.e.\ $[S_x, S_y]\neq 0, [S_x, S_z]\neq 0, [S_y, S_z]\neq0$---they represent incompatible observables that cannot be simultaneously measured. Let us then consider the following scenario: an electron is prepared with spin up along the $x$ direction and subsequently the spin along the $y$ axis is measured. Let the propositions $p$, $q$ and $r$ be defined as follows:
\begin{description}
	\item $p$: ``the electron has $x$-spin up'', 
	\item $q$: ``the electron has $y$-spin up'',
	\item $r$: ``the electron has $y$-spin down''.
\end{description}
As we said, there are only two states in which a particle can be found along a given axis. Therefore the proposition $q \vee r$ ``must be true'' (\cite{Giuntini:2002}, pp.\ 133-134). However, because the different directions are incompatible, if $p$ is true then both $q$ and $r$ must be false. The result is, as noted before, that the quantum disjunction is true even if neither of its members is true.

This behavior in turn breaks the law of distributivity which would state:
\begin{align}
	\label{DL}
	p\wedge(q\vee r)\longleftrightarrow(p\wedge q)\vee(p\wedge r).
\end{align}
\noindent Since we assume that $p$ is true and we know that the disjunction $(q\vee r)$ is also true, we deduce that $p\wedge(q\vee r)=\top$; however, $(p\wedge q)$ and $(p\wedge r)$ are both false in virtue of the incompatibility between the spin operators among different axes. This fact implies that the right hand side of the distributive law asserts a false statement, i.e.\ $(p\wedge q)\vee(p\wedge r)=\bot$. Consequently, we have 
\begin{align}
	\label{contradiction}
	p\wedge(q\vee r)=\top\longleftrightarrow(p\wedge q)\vee(p\wedge r)=\bot. 
\end{align}
Therefore the distributive law fails in the context of quantum theory.\footnote{In the context of quantum logic, it is usually believed that distributivity must be replaced by a weaker law: $(p\wedge q)\vee(p\wedge r)\longrightarrow(p\wedge(q\vee r)).$}

@@

Their analysis begins by defining quantum logical propositions as \emph{experimental propositions}, i.e.\ statements affirming that a certain observable $A$ (or a set of observables) measured on a quantum system $s$ has a given value $a_i$ (or a sequence of values). More precisely, the authors stress that both in classical and quantum mechanics observations of physical systems are given by the readings of the experimental outcomes ($x_1, \dots, x_n$) of \emph{compatible} measurements ($\mu_1, \dots, \mu_n$). The values ($x_1, \dots, x_n$) are elements of what the authors called the ($x_1, \dots, x_n$)-space, i.e.\ the ``observation-space'' of the system in question, whose elements are all the possible combinations of results of the compatible measurements ($\mu_1, \dots, \mu_n$). Hence, the actual values ($x_1, \dots, x_n$) form a subset of such a space. Birkhoff and von Neumann, then, defined the experimental propositions concerning a physical system as the subsets of the observation space associated with it.

Secondly, the authors underlined that in classical and quantum mechanics the states of physical systems are mathematically represented by points in their state spaces---phase space for the classical case, and Hilbert space $\mathcal{H}$ for the quantum case---which provide the maximal information concerning the system. A point in phase space corresponds to the specification of the position and momentum variables of a certain classical system, whereas in QM the points of $\mathcal{H}$ correspond to wave functions. The authors, then, find a connection between subsets of the observation-space of a system and subsets of its Hilbert space, specifying that quantum experimental propositions are mathematically represented by a closed linear subspace of $\mathcal{H}$; this step is crucial in order to obtain the quantum propositional calculus.\ Alternatively, we can say that quantum mechanical operators correspond to propositions with ``yes/no'' (``true/false'') outcome in a logical system\footnote{Cf.\ also \cite{David:2015}, p.\ 78, where we read that ``An orthogonal projector $\textbf{P}$ onto a linear subspace $P\subset\mathcal{H}$ is indeed the operator associated to an observable that can take only the values 1 (and always 1 if the state $\psi\in P$ is in the subspace $P$) or 0 (and always 0 if the state $\psi\in P^{\perp}$ belongs to the orthogonal subspace to $P$).\ Thus we can consider that measuring the observable $\textbf{P}$ is equivalent to perform a test on the system, or to check the validity of a logical proposition $p$ on the system''.} as underlined by Svozi:
\begin{quote}
Any closed linear subspace of---or, equivalently, any projection operator on---a Hilbert space corresponds to an elementary
proposition. The elementary \emph{true--false} proposition can in English be spelled out explicitly as
\begin{quote}
``The physical system has a property corresponding to the associated closed linear subspace'' (\cite{Svozi:1999}, p.\ 1).
\end{quote}
\end{quote}

@@

In this section we will review the examples given in the previous one and explain that the failure of distributivity stems from the ambiguity of the quantum logical propositions and, once that is clarified, standard classical rules can be applied also in the quantum context. More precisely, we are going to argue that (i) it is not necessarily the case that a quantum disjunction must be true when neither of its members is, and (ii) that \eqref{DL} can be retained in QM, since the arguments seen above---usually meant to show its failure---are flawed.

@@

As noted in the previous section, in classical mechanics one generally considers physical systems as objects having well-defined values for inherent, dynamical and relational properties which do not depend on any observation---meaning that measuring a magnitude on a classical particle will reveal a pre-existing value of the observed quantity. Thus, statements about the ontic state of classical systems before or after a measurement can be conflated without any dangerous metaphysical consequence.\footnote{Alternatively stated, in classical mechanics it is usually claimed that the sentence ``The value of the quantity $y$ for the system $s$ is $x$'' can be considered equivalent to the statement ``After a measurement $M$ of the quantity $y$ on the system $s$, the value $x$ is obtained''. This equivalence is due to the fact that measurements in the classical context do not alter the observed system, or in less idealized scenarios, that the disturbance caused by the interaction between observed system and measuring device is negligible. A more elaborate treatment of measurements in classical mechanics goes beyond the scope of the present paper.}

However, in the context of QM this move is not allowed: the effect of a measurement on the state of a quantum system, however one may choose to interpret it, cannot be disregarded. A statement about a system before the performance of a certain measurement---i.e.\ about its preparation\footnote{In operational quantum theory the preparations are a set of instructions that an agent has to follow in order to prepare a quantum system in a certain state. We are aware that preparations can be considered special kinds of measurements. For the sake of the argument, however, in what follows we take preparations to be the instructions to arrange a physical system in a given state \emph{before} the measurement of a certain quantity is actually performed.}---is different from a statement regarding the same object when the measurement has been actually carried out, i.e.\ about the obtained measurement outcome. Indeed, it is only through repeated observations that we can infer anything about the state of the system before the measurement. Thus, propositions like ``the electron has $y$-spin up'', ``the velocity of the particle is $v$'', ``the position of the particle is $x$'' are inherently ambiguous as they do not spell out whether they are concerned with preparations or measurement outcomes. Once such vagueness is taken into account and clarified, the purported failure of classical logic disappears.

In order to support this claim, let us redefine the propositions encountered in the previous example as follows:
\begin{description}
	\item $p_i$: ``before the measurement, the electron $x$-spin is up'', 
	\item $p_o$: ``after the measurement, the electron $x$-spin is up'', 
	\item $q_i$: ``before the measurement, the electron $y$-spin is up'',
	\item $q_o$: ``after the measurement, the electron $y$-spin is up'',
	\item $r_i$: ``before the measurement, the electron $y$-spin is down'',
	\item $r_o$: ``after the measurement, the electron $y$-spin is down''.
\end{description}
\noindent Given the setup of the physical situation at hand, after the measurement of $S_y$ either $q_o$ or $r_o$ must be true, which means $q_o \vee r_o = \top$. This reflects the simple fact that \emph{after} a measurement of spin along the $y$ axis the state of the quantum particle must be one of the two indicated in either $q_o$ or $r_o$. Nonetheless, it is worth noting that $q_i$ and $r_i$ are \emph{not} the same statements as $q_o$ and $r_o$---i.e.\ $q_i$ and $r_i$ are not equivalent to $q_o$ and $r_o$---therefore, it is not necessarily true that $q_o \vee r_o = q_i \vee r_i$. This unwarranted equivalence is what causes the problem, to make evident why it is so, let us add a third step to our example.

After the measurement of $S_y$, in the case we obtained $y$-spin down we will rotate the direction of spin by 90 degrees using the precession generated by a magnetic field, so that $x$-spin will be up, as we can see from Figure 1 below. However, in the other case, when $y$-spin up is obtained, we leave the direction of spin unchanged.

\begin{figure}
\tikzstyle{vecArrow} = [thick, decoration={markings,mark=at position
	1 with {\arrow[semithick]{open triangle 60}}},
double distance=1.4pt, shorten >= 5.5pt,
preaction = {decorate},
postaction = {draw,line width=1.4pt, white,shorten >= 4.5pt}]
\tikzstyle{innerWhite} = [semithick, white,line width=1.4pt, shorten >= 4.5pt]

@@

		%	\draw \setA;
		%	\begin{scope}
		%		\clip \setA;
		%		\draw (-5, 4) .. controls (0, 2) and (0, -2) .. (-5, -4);
		%		\draw (2, 4) .. controls (2, 2) and (0, 1) .. (-1.46, 1);
		%		\draw (1, 1.9) .. controls (1.5, 1) and (2, -2.5) .. (0.5, -4);
		%		\draw (1.49, -1)  .. controls (2, -0.3) and (4, 0.5) ..  (5, 0);
		%	\end{scope}
		%	\node at (-4, -4.5) {$A$};
		%	\node at (-3.2, 0) {$P_1$};
		%	\node at (-0.4, 2.6) {$P_2$};
		%	\node at (-0.2, -1.2) {$P_3$};
		%	\node at (3, -1.7) {$P_4$};
		%	\node at (3, 1.5) {$P_5$};
           \end{tikzpicture} 
}
\end{center}
\caption{\small{Graphic representation of the discussed example. Step 1 describes the preparation of the electron in the $x$-spin up state; Step 2 represents the measurement of $S_y$ at a later time; finally in Step 3 we rotate the direction of spin by 90 degrees if at Step 2 the particle is found in the $y$-spin down state. We leave the direction of spin unchanged otherwise.}}
\end{figure}

Consider now the statements:
\begin{description}
	\item $p_f$: ``after the third step, the electron $x$-spin is up'',
	\item $q_f$: ``after the third step, the electron $y$-spin is up'',
	\item $r_f$: ``after the third step, the electron $y$-spin is down'',
\end{description}
\noindent that concern the state of the electron at the later time, after the possible rotation of spin. By construction, the particle will either have $x$-spin up or $y$-spin up, either $p_f$ or $q_f$ must be true and therefore $p_f \vee q_f = \top$.\footnote{In fact, note that $q_f$ is true if and only if $q_o$ is true, and, similarly, $p_f$ is true if and only if $r_o$ is true.} But again, this relationship is valid only after the third step, which means that $p_o \vee q_o$ is not necessarily true because these latter statements are evaluated at different (earlier) times.
